---
title: "FMLE quickstart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FMLE quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

FMLE fits a fuzzy mixture of linear experts with shared fuzzy gates and expert-specific linear regressors.
This vignette shows a minimal end-to-end workflow on the bundled demo data.

## Load demo data

```{r}
library(FMLE)

demo <- readRDS(system.file("extdata", "fmle_demo.rds", package = "FMLE"))

X_train <- demo$X_train
X_test  <- demo$X_test
Y_train <- demo$Y_train
Y_test  <- demo$Y_test
Z_train <- demo$Z_train
Z_test  <- demo$Z_test

dim(X_train)
dim(Y_train)
dim(Z_train)
```

## Preprocessing 

```{r}
q <- 0.995

cap_and_scale_fit_local <- function(y, q = 0.995, eps = 1e-8) {
  cap <- as.numeric(stats::quantile(y, probs = q, na.rm = TRUE))
  y_cap <- pmin(y, cap)
  y_log <- log1p(y_cap + eps)
  mu <- mean(y_log, na.rm = TRUE)
  sd <- stats::sd(y_log, na.rm = TRUE)
  if (is.na(sd) || sd == 0) sd <- 1
  list(cap = cap, mu = mu, sd = sd, eps = eps)
}

cap_and_scale_apply_local <- function(y, tf) {
  y_cap <- pmin(y, tf$cap)
  y_log <- log1p(y_cap + tf$eps)
  (y_log - tf$mu) / tf$sd
}

tf_y <- cap_and_scale_fit_local(Y_train[, 1], q = q)
y_train <- cap_and_scale_apply_local(Y_train[, 1], tf_y)
y_test  <- cap_and_scale_apply_local(Y_test[, 1], tf_y)
```

## Single-protein quickstart

## Step 1: tune hyperparameters by cross-validation

```{r}
protein_name <- colnames(Y_train)[1]
protein_name
cv <- fmle_cv_parallel(
  X = X_train,
  y = Y_train[, 1],
  Z = Z_train,
  R_grid = c(2, 3),
  m_grid = c(1.6, 1.8),
  lambda_grid = c(0, 1e-3),
  folds = 3,
  seed = 1,
  exec = "sequential",
  verbose = FALSE
)

cv$best
head(cv$table)
```

## Step 2: refit on the full training set using the best selected values

```{r}
best <- cv$best

fit <- fmle_train(
  X = X_train,
  y = y_train,
  Z = Z_train,
  R = best$R,
  m = best$m,
  lambda_l1 = best$lambda,
  ridge = 1e-6,
  standardize = TRUE,
  seed = 1
)

class(fit)
fit$R
fit$m
```

## Step 3: predict on held-out cells

```{r}
pred <- fmle_predict(
  model = fit,
  X_new = X_test,
  Z_new = Z_test,
  return_se = TRUE
)
```

## Step 4: evaluate predictions

```{r}
pearson <- cor(pred$mean, y_test, method = "pearson")
spearman <- cor(pred$mean, y_test, method = "spearman")
mse <- mean((pred$mean - y_test)^2)

data.frame(
  metric = c("Pearson", "Spearman", "MSE"),
  value = c(pearson, spearman, mse)
)
```

## Single-task results across multiple proteins

```{r}
proteins_to_show <- colnames(Y_train)
res_list <- vector("list", length(proteins_to_show))

for (j in seq_along(proteins_to_show)) {
  prot <- proteins_to_show[j]

  tf_y <- cap_and_scale_fit_local(Y_train[, j], q = q)
  y_train_j <- cap_and_scale_apply_local(Y_train[, j], tf_y)
  y_test_j  <- cap_and_scale_apply_local(Y_test[, j], tf_y)

  cv_j <- fmle_cv_parallel(
    X = X_train,
    y = Y_train[, j],
    Z = Z_train,
    R_grid = c(2, 3),
    m_grid = c(1.6, 1.8),
    lambda_grid = c(0, 1e-3),
    folds = 3,
    seed = 1,
    exec = "sequential",
    verbose = FALSE
  )

  best_j <- cv_j$best

  fit_j <- fmle_train(
    X = X_train,
    y = y_train_j,
    Z = Z_train,
    R = best_j$R,
    m = best_j$m,
    lambda_l1 = best_j$lambda,
    ridge = 1e-6,
    standardize = TRUE,
    seed = 1
  )

  pred_j <- fmle_predict(
    model = fit_j,
    X_new = X_test,
    Z_new = Z_test,
    return_se = TRUE
  )

  res_list[[j]] <- data.frame(
    protein = prot,
    R = best_j$R,
    m = best_j$m,
    lambda = best_j$lambda,
    Pearson = cor(pred_j$mean, y_test_j, method = "pearson"),
    Spearman = cor(pred_j$mean, y_test_j, method = "spearman"),
    MSE = mean((pred_j$mean - y_test_j)^2)
  )
}

res_tab <- do.call(rbind, res_list)
res_tab$Pearson <- round(res_tab$Pearson, 3)
res_tab$Spearman <- round(res_tab$Spearman, 3)
res_tab$MSE <- round(res_tab$MSE, 3)
res_tab
```

## Important preprocessing note

For single-task FMLE, fmle_cv_parallel() internally applies cap/log/scale preprocessing to the response before fold-wise fitting and evaluation. In contrast, fmle_train() fits the response exactly as supplied.

After selecting (R, m, lambda) by cross-validation, refit the full model using the response scale you intend to use for the final model and evaluation.



